---
title: "The options of *fclust_plot*"
author: "Benoît Jaillard"
date: "`r Sys.Date()`"
output: 
  knitr:::html_vignette:
    toc: true
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{The options of fclust_plot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  out.width = "100%"
)
```


The function *fclust_plot* can conveniently plot
the primary and secondary trees of component clustering,
observed, modelled and predicted performances of assemblages,
composition and mean performances of assembly motifs,
mean performances of assemblages containing a given components,
or performances of some given assemblages.


```{r init_plotfclust, echo=FALSE, results="hide"}
library(functClust)
nbElt <- 16
dat.2004 <- CedarCreek.2004.2006.dat[ , c(1, 1 + 1:nbElt, 1 + nbElt + 1)]
```



## The option *opt.tree*

**opt.tree** manages the plot of primary and secondary trees
of component clustering. The option appears as a list 
of sub-options : 
*opt.tree = list("cal", "prd", "leg", cols, "zoom", window, "all")*.

* **"cal"** plots the primary tree of component clustering.
   A blue dashed line indicates the optimum number of component clusters.
   A red dashed line indicates the value of tree efficiency.

* **"prd"** plots the validated, secondary tree of component clustering.
   A red dashed line indicates the value of tree efficiency.

* **"leg"** : if **"cal"** or **"prd"** are checked, the option **leg** 
  plots the content in components of validated clusters 
  of the secondary tree.

* **cols** gives *a priori* defined colours to each component.
   This option is useful to identify some given components,
   or components *a priori* clustered, into an *a posteriori* clustering.

* **"zoom"** : if **"cal"** or **"prd"** are checked, 
  the option **"zoom"** allows to only plot the first, 
  significant component clusters. 
  **window = n** (n = number of components to plot) must then be informed.

* **"all"** is equivalent to *opt.tree = list("cal", "prd", "leg")*.


```{r plot_opt_tree, results="hide", out.width="48%", fig.align="default", fig.show="hold", fig.cap=c("Opt.tree manages the plot of primary and secondary trees")}
apriori <- c("F","C3","L","C4","F","C3","C3","L","F","L","F","C4","L","C3","C4","C4")
res <- fclust(dat.2004, nbElt, opt.mod = "byelt", opt.mean = "gmean")

fclust_plot(res, main = "BioDiv2", opt.tree = list("prd"))
fclust_plot(res, main = "BioDiv2", opt.tree = list("prd", cols = apriori))
```

Both the graphs show the validated, secondary tree 
pruned above the optimum number of component clusters. 
However, on left graph components belonging to a cluster are of same colour,
on right graph components belonging to clusters *a priori* defined
by the user are of same colour. 
Here, the meadow species are *a priori* clustered
in Legumes (in red), Forbs (in cyan),
C3-grasses (in blue) and C4 grasses (in gold).
The *x-axis* of tree is the system components.
The *y-axis* of tree is the quality criterion of modelling, 
*i.e.* the coefficient of determination *R2*. 
Modelling efficiency (*E-value*) is indicated by a red dashed line.
Values of *R2*, *E* and *E/R2* are indicated. 


## The option *opt.perf*

**opt.perf** manages the plot
of observed, modelled and predicted performances of assemblages,
and associated statistics. The option appears as a list 
of sub-options : *opt.perf = list("stats_I", "stats_II", "cal", "prd", "pub", "missing", "calprd", "seq", "ass", "aov", pvalue, "all")*.

* **"stats_I"**, **"stats_II"** plot the statistics associated to
   fit of primary and secondary trees.


```{r plot_opt_perf_stats, results="hide", out.width="48%", fig.align="default", fig.show="hold", fig.cap=c("Opt.perf=list('stats_xx') manages the plot of statistics")}
fclust_plot(res, main = "BioDiv2", opt.perf = list("stats_II"))
```

The first left graph shows coefficient of determination *R2* (black circles) 
and efficiency *E* (red circles) of validated secondary. 
The right graph shows the proportion of assemblages of which the performance
cannot be predicted by cross-validation only using the clustering model
(blue squares).
Here all performances are predicted by using the whole secondary tree.
Both the lowest graphs show AIC (blue circles) 
and AICc (red circles) of models.
Green dashed line indicates the optimum number of component clusters.



* **"cal"**, **"prd"**, **"pub"** : 
    **"cal"** plots modelled performance *versus* observed performance, 
    **"prd"** plots performances modelled and predicted by cross-validation 
        *versus* observed performance,
    **"pub"** plots performances modelled and predicted by cross-validation 
    *versus* observed performance using only one colour and symbol, 
    which is useful for publication.


```{r plot_opt_perf_prd, results="hide", out.width="48%", fig.align="default", fig.show="hold", fig.cap=c("Opt.perf=list('prd') on left, =list('pub') on right")}
fclust_plot(res, main = "BioDiv2", nbcl = 3, opt.perf = list("prd", "aov", pvalue = 0.01, "pub"))
```

Both graphs show the model goodness-of-fit,
left and right graphs differing only by the used colours and symbols.
On left graph, the performances of assemblages
belonging to different assembly motifs
are plotted with different colours and symbols.
The symbols correspond to performances predicted by the model 
(modelled performances), of which the goodness-of-fit 
is evaluated by tree coefficient of determination *R2*.
The vertical lines correspond to performances 
independently predicted by cross-validation
(independently predicted performances),
of which the quality is evaluated by tree efficiency *E*.
Global statistic are indicated (number of component clusters, 
number of assembly motifs, *R2*,
*E* and the ratio *E/R2*).
A variance analysis is indicated on detailled graph,
on the left side for modelled performances, 
on the right side for performances predicted by cross-validation,
using a *pvalue = 0.01*. 
On both graphs, the solid red line is line 1:1, 
and the blue dashed lines are mean observed and modelled performances.



* **"missing"** identifies the assemblages
    of which the performance cannot be predicted
    using the clustering model of the current level.

* **"calprd"** plots performances predicted by cross-validation
    *versus* performances predicted by clustering model
    ("modelled performances").


```{r plot_opt_perf_missing, results="hide", out.width="48%", fig.align="default", fig.show="hold", fig.cap=c("Opt.perf=list('missing') on left, =list('calprd') on right")}
fclust_plot(res, main = "BioDiv2", nbcl = 4, opt.perf = list("missing"))
fclust_plot(res, main = "BioDiv2", nbcl = 4, opt.perf = list("calprd", "aov", pvalue = 0.01))
```

On the left graph, we observe that 
the performances of all assemblages can be predicted 
using *nbcl = 4* clusters of components, except the assemblages
"110",  "211" and  "237". On the right graph, we observe 
that the performances of all assemblages are as well predicted as modelled,
except those of 3 assemblages (2 plotted as red circle, and one as gold square).

The right graph shows predicted performances *versus* modelled performances.



* **"seq"** plots performances of assembly motifs,
   from 1 to the optimum number of component clusters.

```{r plot_opt_perf_seq, results="hide", out.width="48%", fig.align="default", fig.show="hold", fig.cap=c("Opt.perf=list('prd', 'seq') plots performances predicted from1 to nbOpt component clusters")}
fclust_plot(res, main = "BioDiv2", opt.perf = list("prd", "seq", "aov", pvalue = 0.01))
```

The number *m* of assembly motifs generates 
by the combination of *nbcl* component clusters is *m = 2^nbcl - 1*. 
It is 1, 3, 7, 15 and 31 for 1, 2, 3, 4 and 5 clusters.
Here the optimum number of clusters is 4, and 13 among 15 assembly motifs
are observed in the experiment.


* **"ass"** plots the name of each assemblage
    close to its performance. 


```{r plot_opt_perf_ass, results="hide", out.width="48%", fig.align="default", fig.show="hold", fig.cap=c("Opt.perf=list('prd', 'ass') adds the names of assemblages on plot")}
fclust_plot(res, main = "BioDiv2", opt.perf = list("prd", "aov", pvalue = 0.01))
fclust_plot(res, main = "BioDiv2", opt.perf = list("prd", "ass"))
```


* **"aov"** does a variance analysis of assemblage performances
    by assembly motifs, and plot the results. **pvalue = val** 
    (val = threshold of probability) must then be informed. If not,
   *pvalue = 0.05* by default.

* **"all"** is equivalent to
   *opt.perf = list("cal", "prd", "pub", "calprd", "aov", pvalue = 0.05)*.


## The option *opt.ass*

**opt.ass** plot different performances of some given, identified assemblages.
The option is activated only when an assemblage performance 
is observed over several experiments 
(see vignette **The option *xpr* : multi-functionality in *functClust* **). 


## The option *opt.motif*

**opt.motif** manages the plot of mean performances
 of assembly motifs as boxplots.
 The option is a list, that can include
 *opt.motif = list("obs", "cal", "prd", cols, "hor", "seq",
 pvalue, "all")*:

* **"obs", "cal", "prd"**  plot the observed,
    modelled or predicted by cross-validation mean performances
    of assembly motifs as boxplots.
    Assembly motifs are named as the combinations of component clusters.

* **"hor"** plot boxplots as horizontal boxes, 
   that is x-axis corresponds to assemblage performances,
   and y-axis corresponds to assembly motifs.

* **"seq"** plots mean performances of assembly motifs,
   from 2 to the optimum number of component clusters.

* **"aov"** does a variance analysis of assemblage performances
    by assembly motifs, and plot the results. **pvalue = val** 
    (val = threshold of probability) must then be informed. If not,
   *pvalue = 0.05* by default.

* **"all"**  is equivalent to
   *opt.motif = list("obs", "cal", "prd", "seq", "aov", pvalue = 0.05)*.


```{r plot_opt_motif, results="hide", out.width="48%", fig.align="default", fig.show="hold", fig.cap=c("Opt.motif=list('obs', 'seq') plots performances sorted by assembly motif")}
fclust_plot(res, main = "BioDiv2", opt.motif = list("prd", "hor", "seq", "aov", pvalue = 0.01))
```

The boxplots show observed performances of assemblages 
belonging to different assembly motifs.
Means are indicated with different symbols, 
of same shape and colour as those used in *opt.perf*.
The size (in number of assemblages) of each assembly motifs
is indicated on the left side of the graph, and
the results of a variance analysis of observed performances
by assembly motif is indicated on the right side of the graph. 
The blue dashed lines are mean modelled performances (opt.motif = "prd")
and red dashed lines are mean observed performances.



## The option *opt.comp*

**opt.comp** manages the plot as boxplot
 of observed mean performances
 of assemblages that contain a given component.
 The option can include
 *opt.comp = list("tree", "perf", "hor", cols,
 pvalue, "zoom", window, "all")*:
 
* **"tree", "perf"** plot the observed mean performances
    of assemblages that contain a given component as boxplots,
    sorted like the *clustering tree*
    or sorted by *increasing mean performance*.

* **"hor"** plot boxplots as horizontal boxes, 
   that is x-axis corresponds to assemblage performances,
   and y-axis corresponds to assembly motifs.

* **cols** allows to identify specified components,
   for instance *a priori* clustering 
   into an *a posteriori* clustering.

* **"aov"** does a comparison od means between
   performance of set of assemblages that contain a given component 
   against performance of set of other assemblages 
   that do not contain the given component in consideration. 
   Each compoent is analised separately. 
   The result is *"a"* if performance of assemblage set is significantly higher, 
   *"b"* if performance of assemblage set is not different, 
   and *"c"* if performance of assemblage set is significantly lower,
   than performance of set of other assemblages 
   that do not contain the given component.
   **pvalue = val**(val = threshold of probability) must then be informed.
   If not, *pvalue = 0.05* by default.

* **"all"**  is equivalent to
   *opt.comp = list("tree", "aov", pvalue = 0.05, "zoom", window = 20)*.


```{r plot_opt_comp, results="hide", out.width="48%", fig.align="default", fig.show="hold", fig.cap=c("Opt.comp=list('tree', 'leg') plots set of performances that contain a given component, sorted as the hierarchical tree of components")}
fclust_plot(res, main = "BioDiv2", opt.comp = list("tree", "leg", "aov", pvalue = 0.01))
```

The boxplots observed mean performances
of assemblages that contain a given component,
sorted as the *clustering tree*.
It is why we prefer a vertical rather than a horizontal plot.
The size (in number of assemblages) of each set
is indicated on the left side of the graph.
The results of a variance analysis of observed performances
is indicated on the right side of the graph. 
The blue dashed lines are mean modelled performances (opt.comp = "prd")
and red dashed lines are mean observed performances.



