---
title: "Testing significance of different variables with *functClust*"
author: "Benoît Jaillard"
date: "`r Sys.Date()`"
output: 
  knitr:::html_vignette:
    toc: true
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Testing significance with functClust}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  out.width = "100%"
)
```



```{r init_testfclust, echo=FALSE, results="hide"}
library(functClust)
fres <- CedarCreek.2004.2006.res
```



## Are all components or observations equally important?

Functional clustering aims to identify the role played 
by each component belonging to an interactive system 
on the genesis of a collective, system-specific performance.
It needs the observation of a collection of assemblages of components
of different elemental composition,
of which collective, system-specific performances are observed.
A functional clustering groups the components of the interactive system
on the basis of their effects on the system-specific performance.
The method hierarchises *de facto* the functional groups:
the first division explains most of the observed variance, 
and the subsequent divisions explain a smaller and smaller part 
of the observed variance. 

Functional clustering thus leads to ask the following questions: 
are some components that belong to each functional group 
more efficient than others? 
How can component efficiency be assessed? 
Is it possible to priorize components 
for their effect on system performance?
Similar questions can be raised about relative importance of 
different observations, that is different observed component assemblages 
of the collection, or of different observed performances in the case of 
repeated observations of the performance, or the observations of different properties carried out by the interactive system.

The function *ftest* answer to the questions.
The method is based on removing one element of the dataset 
and evaluating the perturbation induced on functional clustering.
The removed element can be a system component, a component assemblage,
or an assemblage performance. The induced perturbation is evaluated 
by comparing the clustering tree obtained with removing the element,
to the reference clustering tree obtained with the complete dataset.
Different indices are computed, using the R-package **clusterCrit**
(Package "clusterCrit": Clustering Indices, by Bernard Desgraupes, 
University of Paris Ouest - Lab Modal'X). The indices are: 
*Precision* index and *Recall* index, 
or indices proposed by different authors as *Czekanowski_Dice* index, *Folkes_Mallows* index, *Jaccard* index, *Kulczynski* index,
*Rogers_Tanimoto* index, *Russel_Rao* index, and 
both *Sokal_Sneath1* and *Sokal_Sneath2* indices.



## The function *ftest* and its options

The function *ftest* needs first that the function *fclust* is run 
for computing the reference clustering tree obtained 
with the complete dataset. 
The object returned by *fclust* is noted *fres*. 
The options of *ftest* are:

* **opt.var** determines the variable to treat. 
The option can be **"components"**, **"assemblages"** or 
**"performances"**. 

* **opt.nbMax** determines the tree level to be reached. 
The function *ftest* is very time-consuming 
since it is based on the repeat of functional clusterings. 
Then, only the validated lowest part of the tree is interesting,
from trunk (*tree-level = 1*) until the optimum level *fres\$nbOpt*.
Beyond the optimum level *fres\$nbOpt*, the clustering 
does not bring any supplementary information.
The computation therefore deserves to be stopped at this tree level.
In *ftest*, the option *opt.nbMax = fres\$nbOpt* by default. 
The option *opt.nbMax* allows the user to change the stop tree-level,
more or less.

* **opt.R2** determines if the user needs to compute other statistiques
than clustering indices. If *opt.R2 = TRUE*, each primary tree is validated
and the vectors of coefficient of determination *R2*
and efficiency *E* are computed.

* **opt.plot** determines if the user wish to follow the computation 
by plotting the resulting tree at each element removing.
If option *opt.plot = TRUE*, the resulting tree is plotted 
at each element removing. 
The removed element is indicated on the tree plot.

The function *ftest* returns an object *rtest*, which consists on 
a list of matrices, each matrix containing
the results for a given clustering index, 
and *R2* and *E* if *opt.R2* was checked.



## The function *ftest_plot*

The function *ftest_plot* first needs 
the object *fres* generated by the function *fclust*,
and the object *rtest* generated by the function *ftest*.
The options of *ftest_plot* are:

* **opt.crit** determines the indices to be plotted. 
The option *opt.crit* is a list or a vector of characters
choisen among **"Czekanowski_Dice"**, **"Folkes_Mallows"**, **"Jaccard"**,
**"Kulczynski"**, **"Precision"**, **"Rand"**, **"Recall"**,
**"Rogers_Tanimoto"**,
**"Russel_Rao"**, **"Sokal_Sneath1"** and **"Sokal_Sneath2"**.

* **opt.var** determines the variable to plot. 
The option can be **"components"**, **"assemblages"** or 
**"performances"**. 



```{r optplottestcomp, results="hide", out.width="48%", fig.align="default", fig.show="hold", fig.cap=c("Raw tree on left, Tree with components sorted by decreasing effect on right")}
fclust_plot(fres = CedarCreek.2004.2006.res, opt.tree = "prd")

ftest_plot(fres = CedarCreek.2004.2006.res, 
           rtest = CedarCreek.2004.2006.test.components,
           main = "BioDIV2", 
           opt.var = "comp", opt.crit = "Jaccard", opt.comp = "sorted.tree")
```

The graph on the left is the raw tree, 
directly  obtained with the function *fclust*. 
On the right, the components of the tree are sorted 
by their decreasing effect of *Jaccard* index when they are one by one
removed from dataset.
For instance, within the component cluster *"b"* (in blue), 
the effect induced by species on ecosystem biomass can be sorted as: 
*"Liass"* > *"Lesca"* > *"Amocan"*.



```{r optplottestass, results="hide", out.width="48%", fig.align="default", fig.show="hold", fig.cap=c("Raw tree on left, Tree with components sorted by decreasing effect on right")}
ftest_plot(fres = CedarCreek.2004.2006.res, 
           rtest = CedarCreek.2004.2006.test.assemblages, 
           main = "BioDIV2", opt.var = "assemblages")
```

The graph shows the mean *Jaccard* index when each assemblage is one by one
removed from dataset. And the text indicates the assemblages 
sorted by decreasing effect within each assembly motif 
on the functional clustering.  
For instance, the assembly motif *ad* has the highest mean performance.
The effect induced by removing assemblage can be sorted as: 
*"plot 193"* > *"234"* > *"300"* > *"342"*.



## The functions *ftest_write* and *ftest_read*

The functions **ftest_write** and **ftest_read** allow 
to save and load, respectively, the results 
generated by the function *ftest*.



## Note 

Note that some computations are time-consuming. 
To facilitate the monitoring of the smooth running of the computations,
informations are written on the Console 
and graphs are drawn on the Plots panel. 
The writting are enable or disable by the "verbose" option. 

```{r foptions_6}
getOption("verbose")
# to follow the computations
options(verbose = TRUE)
# to deactivate the option
options(verbose = FALSE)
```

