---
title: "The options of *fclust*"
author: "Benoît Jaillard"
date: "`r Sys.Date()`"
output: 
  knitr:::html_vignette:
    toc: true
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{The options of fclust}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  out.width = "100%"
)
```


The function *fclust* can work using different methods and models 
managed by options.
The three main options are **opt.method**,
**opt.model** and **opt.mean**. 
Additional option is **opt.jack**.
The options are defined by default
for focusing on the main results provided by a functional clustering.
They are:  *opt.tree = list("prd", "leg"), opt.perf = list("prd", "pub"), opt.motif = list("obs", "hor", "leg")*.



```{r init_opt_fclust, echo=FALSE, results="hide"}
library(functClust)
nbElt <- 16
dat.2004 <- CedarCreek.2004.2006.dat[ , c(1:(nbElt + 2))]
```


## The option *opt.method*

**opt.method** determines the method of clustering.
The option can be **"divisive"**, **"agglomerative"** or **"apriori"**.
All methods generate hierarchical trees.
Each tree is complete, running from a unique trunk
to as many leaves as components.

* If **opt.method = "divisive"**, the components are clustered
 from the trivial cluster where all components are together,
 towards the clustering where each component is isolated in a cluster.
 This method is very efficient because the first component partioning
 change considerably the coefficient of determination of clustering model.
 The first divisions are thus very discriminative, and
 it is therefore possible to surely observe the effect on performance
 of co-occurring components within assemblages.

* If **opt.method = "agglomerative"**, the components are clustered
 from the trivial clustering where each component is isolated in a cluster,
 towards the cluster where all components are together. 
 In most cases, the first component clustering do not change
 the coefficient of determination of clustering model: 
 the first clustering are not discriminative and
 it is therefore not possible to select the component combinations 
 associated with the strongest effects of assemblage performance.
 
 
```{r opt_method, results="hide", out.width="48%", fig.align="default", fig.show="hold", fig.cap=c("Opt.method determines the method of clustering: on left 'divisive', on right 'agglomerative'")}
res <- fclust(dat.2004, nbElt, opt.method = "divisive")
fclust_plot(res, opt.tree = list("prd"))

res <- fclust(dat.2004, nbElt, opt.method = "agglomerative")
fclust_plot(res, opt.tree = list("prd"))
```

The left graph shows the tree obtained using *"divisive"*" method,
the right graph the tree obtained using *"agglomerative"*" method.
Divisive method gives generally a more accurate and a more predictive tree
than agglomerative method. Divisive and agglomerative methods 
give the same results whether the number of components is small,
thus the number of possible component partitions also is small.
(The possible component partitions is given by the number of Stirling
of second species, see the function *stirling*).

* **opt.method = "apriori"** : the option assumes that the user knows
an *a priori* partitioning of the system components he is studying.
The partition is arbitrary, in any number of clusters of components,
as long as it is complete, *i.e.* it includes as many components 
as the system includes. In this option, an *a priori* component partition, noted **affectElt**, should be provided. 
**affectElt** is a vector of *integers* or *characters* of length *nbElt*.
A hierarchical tree is then built, by using *opt.method = "divisive"*
from the *a priori* defined component partition
towards the uppermost part of tree (towards tree leaves),
and by using *opt.method = "agglomerative"*
from the *a priori* defined component partition
towards the lowest part of tree (towards tree trunk). 
The resulting tree is therefore
forced by the given, *a priori* component partition *affectElt*.  
Note that *opt.method = "apriori"* with *affectElt = rep(1, nbElt)*
is equivalent to *opt.method = "divisive"*, 
since the *a priori* partitioning consists of a single cluster
that brings together all the components of the system. 
In contrary, *opt.method = "apriori"* with *affectElt = seq_len(nbElt)* 
is equivalent to *opt.method = "agglomerative"*, 
since the *a priori* partitioning consists of isolating each component 
in a cluster.


```{r opt_apriori, results="hide", out.width="48%", fig.align="default", fig.show="hold", fig.cap=c("Opt.method='apriori' forces the hierarchical tree to include a given component partition")}
apriori <- c(1,3,2,4,1,3,3,2,1,2,1,4,2,3,4,4) 
apriori <- c("F","C3","L","C4","F","C3","C3","L","F","L","F","C4","L","C3","C4","C4")

res <- fclust(dat.2004, nbElt, opt.method = "apriori", affectElt = apriori)
fclust_plot(res, opt.tree = list("prd", "leg", cols = apriori))
```

In ecology, meadow species are classically *a priori* clustered
in Legumes (in red), Forbs (in cyan),
C3-grasses (in blue) and C4 grasses (in gold). A functional clustering 
suggests that only the legumes group is pertinent. 



## The option *opt.model*

**opt.model** determines the model for predicting assemblage performance.
The option can be **"bymot"** or **"byelt"**.

*  **opt.mod = "bymot"**: the option is the simplest one. 
 The performances are modelled 
 as the mean performance of assemblages that share a same assembly motif,
 by including all assemblages that belong to the same assembly motif.
 Consequently, all assemblages that share a same assembly motif
 have the same predicted performance, whatever their component composition.
 The modelling of assemblage performances is therefore based only 
 on the partitioning of components into functional groups, 
 and the resulting partitioning of assemblages into assembly motifs. 
 It does not take into account the fact 
 that the elementary composition of each assemblage differs 
 from one assemblage to another within the same assembly motif,
 and that these elementary compositions are known.
 
*  **opt.model = "byelt"**: the option uses the fact
 that the elementary composition of each assemblage differs 
 from one assemblage to another within the same assembly motif,
 and that these elementary compositions are known.
 Within each assembly motif, the performances of assemblages 
 that contain a given components are first averaged. 
 Consequently, a mean value of performance is associated 
 with each component that occurs within the assembly motif. 
 Second the performance of each assemblage is computed 
 as the mean of mean performances of assemblages
 that contain the same components as the assemblage to predict.
 If no assemblage contains component belonging to assemblage to predict,
 performance is computed as the mean performance of all assemblages
 that share a same assembly motif, as in *opt.mod = "bymot"*. 
 As a whole, this procedure partitions assemblages by assembly motif 
 and adds a linear model within each assembly motif
 based on the component occurrence within each assemblage.
 This procedure can improve the explanatory and predictive 
 abilities of the modelling 
 (see Jaillard *et al*., 2018a, Meth. Ecol. Evol.).


```{r opt_model, results="hide", out.width="48%", fig.align="default", fig.show="hold", fig.cap=c("Opt.model determines the model for predicting assemblage performance")}
res <- fclust(dat.2004, nbElt, opt.mod = "bymot")
fclust_plot(res, opt.tree = list("prd"), opt.perf = list("prd", "aov", pvalue = 0.01))

res <- fclust(dat.2004, nbElt, opt.mod = "byelt")
fclust_plot(res, opt.tree = list("prd"), opt.perf = list("prd", "aov", pvalue = 0.01))
```

Both the highest graphs correspond to *opt.model = "bymot"*,
both the lowest graphs to *opt.model = "byelt"*.
The resulting trees are different:
the first *red*group contains *Luppe* in both the trees,
the second *blue*-group contains *Liaas* and *Lesca* in both the trees,
but differs by *Amocan* and *Koecr*,
the third *gold"*group contains *Andge* in both the trees, 
but *Koecr* in *bymot*-tree and several other species in *byelt*-tree, etc....
The coefficients of determination are equivalent 
(*R2* = 0.906 against 0.909), and
the predictive ability of assemblage performances are more robust 
with *opt.model = "bymot"* than with *opt.model = "byelt"*
(*E* = 0.851 against 0.797, then *E/R2* = 0.940 against 0.877).
However, our experiment suggests that *opt.model = "byelt"* 
gives the most likely result. 



## The option *opt.mean*

**opt.mean** determines the formula to use in averaging.
The option can be **"amean"** or **"gmean"**.
Functional clustering is based on 
computations of mean performances of assemblages, differently partitioned. 
The mean formula to use depends on the distribution of assemblage performance:
it can shift a little the resuls.

* If **opt.mean = "amean"**, mean performances are computed 
using an arithmetic formula.

* If **opt.mean = "gmean"**, mean performances are computed 
using a geometric formula.


```{r opt_mean, results="hide", out.width="48%", fig.align="default", fig.show="hold", fig.cap=c("Opt.mean determines the formula to use in averaging")}
res <- fclust(dat.2004, nbElt, opt.mean = "amean")
fclust_plot(res, opt.tree = list("prd"), opt.perf = list("prd", "aov", pvalue = 0.01))

res <- fclust(dat.2004, nbElt, opt.mean = "gmean")
fclust_plot(res, opt.tree = list("prd"), opt.perf = list("prd", "aov", pvalue = 0.01))
```

The left graph corresponds to *opt.mean = "amean"*,
the right graph to *opt.mean = "gmean"*.
The resulting trees are the same, and the model goodness-of-fit 
(*R2* = 0.909 against 0.940; 
*E* = 0.797 against 0.798) are not significantly different.


## The option *opt.jack*

**opt.jack** determines the method of cross-validation. 
By default (*opt.jack = FALSE*), the performance of each assemblage
is predicted by a Leave-One-Out method:
the performance of each assemblage is predicted 
as the mean performance of assemblages
that share a same assembly motif,
except the only assemblage to predict.
If the number of assemblages that share a same assembly motif is large,  Leave-One-Out method is time-consuming. 
It is more convenient to switch towards a jackknife method
(*opt.jack = TRUE*):
the performances of assemblages that belong to each subset are predicted 
as the mean performance of assemblages of other subsets,
except the assemblage subset to predict. 
**jack** then specifies how to divide the assemblage collection.
*jack* is an integer vector of length *2*:
the first integer specifies the size of subset,
the second integer specifies the number of subsets.



## Note 

Note that some computations are time-consuming. 
To facilitate the monitoring of the smooth running of the computations,
informations are written on the Console 
and graphs are drawn on the Plots panel. 
The writting are activated or deactivated by the "verbose" option. 

```{r foptions_3}
getOption("verbose")
# to follow the computations
options(verbose = TRUE)
# to deactivate the option
options(verbose = FALSE)
```


