---
title: "Multi-functionality with *functClust*"
author: "Benoît Jaillard"
date: "`r Sys.Date()`"
output: 
  knitr:::html_vignette:
    toc: true
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Multi-functionality with functClust}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  out.width = "100%"
)
```



```{r init_plotfclust, echo=FALSE, results="hide"}
library(functClust)
nbElt <- 16
dat <- CedarCreek.2004.2006.dat
```



## Repeated observations of a given performance or multi-functionality of a system

The inputs of *fclust* can include several observed performances
generated by the same component assemblages. 
The different performances can correspond to 
a same system function or property observed at different times, 
in different places or under different environmental conditions.
It is then repeated observations of a given system function or property.
All observed performances are of same nature : they are equivalent 
and, generally, do not deserve to be differently weighted.
This is for instance the biomass yearly harvested in 2004, 2005 and 2006 
in Biodiversity II experiment done at Cedar Creek 
by David Tilman and his collaborators
("y2004", "y2005" and "y2006" in *CedarCreek.2004.2006.dat*).

The different performances can also correspond to 
different functions or properties generated 
by a same, given collection of component assemblages belonging to a given system.
The observed performances are of different nature, they are not equivalent, 
and each one deserves perhaps to be differently weighted.
It is then about the understanding of multi-functionality of a system.
This would be the case, for instance, 
for biomass of shoots and biomass of roots
harvested for a given year 
in Biodiversity II experiment done at Cedar Creek.

Both the cases can be analysed by the functions *fclust* 
and plotted by the function *fclust_plot*. 
In the following, we term *"experiment"* 
the observation of a performance, whatever the performance, 
on a given collection of component assemblages
belonging to a given system.
If the inputs of *fclust* include several observed performances
generated by the same component assemblages,
the option **xpr** (for e**XP**e**R**iment) is activated 
in the code of *functClust*.


## The option *"weight"* of *fclust*

The function *fclust* needs at least the name of dataset file
and the number *nbElt* of components belonging 
to the interactive system in consideration
(see vignette **b.The simplest use of *functClust* **).
The number **nbXpr** of observed performances is thus known:
*nbXpr = dim(dataset)[2] - (1 + nbElt)*.
If *nbXpr* is higher then 1,
the option *xpr* is activated.


* **weight** : If there are several experiments, 
the option **weight** of function *fclust* must then be filled.
*weight* is a vector of numerics, 
that specifies the weight to associate to each observed performance 
in the computations of both the quality criterion 
*R2* (coefficient of determination of tree) and *E* (efficiency of tree).
If the option *weight* is not filled, 
all observed performances are considered as equivalent
(*weight* is equal to 1 for each experiment)
(see *Reference* for more informations).


```{r init_opt_weight, results="hide", out.width="48%", fig.align="default", fig.show="hold", fig.cap=c("Weight = c(1,1,1) on left, = c(2,3,1) on right tree")}
res <- fclust(dat, nbElt, weight = c(1, 1, 1), opt.mod = "byelt", opt.mean = "gmean")
fclust_plot(res, main = "BioDiv2", opt.tree = list("prd"))

res2 <- fclust(dat, nbElt, weight = c(2, 3, 1), opt.mod = "byelt", opt.mean = "gmean")
fclust_plot(res2, main = "BioDiv2", opt.tree = list("prd"))
```


## The options of *fclust_plot*

The function *fclust* returns un object 
that contains all necessary informations for plotting the results,
especially the names of different experiments.
If the number *nbXpr* of experiments is higher than 1, 
the option *xpr* is activated.
Then, the user can require that the results of functional clustering
are plotted as a whole (*xpr=all*) and experiment by experiment (*xpr=names(xpr)*). That only needs to add the option *multi* 
in the list of options. 
In the following, the changes are noted and illustrated for each option.


* **opt.tree** manages the plot of primary and secondary trees
of component clustering. 
Any functional analysis generates only a primary and a secondary tree. 
The analysis of several performances does not change anything. 


* **opt.perf** manages the plot
of observed, modelled and predicted performances of assemblages
Functional analysis generates only a primary and a secondary tree. 
The analysis of several performances does not change anything. 


```{r plot_opt_xpr_perf, results="hide", out.width="48%", fig.align="default", fig.show="hold", fig.cap=c("Opt.perf = list('xpr') plots the performances over time")}
fclust_plot(res, main = "BioDiv2", opt.perf = list("multi"))
```


* **opt.ass** plots different performances of some given, identified assemblages.
The option is activated only when the option **xpr** is activated,
that is when a same assemblage performance is observed 
over several experiments, at different times, places 
or environmental conditions,
or when different assemblage performances are generated 
by given assemblages belonging to a system.
Consequently, it is not necessary to specify *multi* 
in the list of options.


```{r plot_opt_xpr_ass, results="hide", out.width="48%", fig.align="default", fig.show="hold", fig.cap=c("Opt.ass = list(who = c('20', '138')) plots the performances over time of assemblages '20' and '138' ")}
fclust_plot(res, main = "BioDiv2", opt.ass = list(who = c("20", "138")))
```


* **opt.motif** manages the plot of mean performances
 of assembly motifs as boxplots.
  When the option **xpr** is activated, 
   the options **"obs", "cal", "prd"** 
   plot mean performances as a whole.
   To plot mean performances experiment by experiment,
  it is necessary to add the option *multi* in the list of options.


```{r plot_opt_xpr_motif, results="hide", out.width="48%", fig.align="default", fig.show="hold", fig.cap=c("Opt.motif = list('obs', 'seq') plots performances over time sorted by assembly motif")}
fclust_plot(res, main = "BioDiv2", opt.motif = list("prd", "hor", "multi", "aov", pvalue = 0.01))
```

* **opt.comp** manages the plot as boxplot
 of observed mean performances
 of assemblages that contain a given component.
 When the option **xpr** is activated,
 the options **"tree", "perf"** 
   plot mean performances as a whole.
   To plot mean performances experiment by experiment,
  it is necessary to add the option *multi* in the list of options.


```{r plot_opt_xpr_comp, results="hide", out.width="48%", fig.align="default", fig.show="hold", fig.cap=c("Opt.comp = list('tree') plots performances over time sorted by component clusters as in hierarchical tree")}
fclust_plot(res, main = "BioDiv2", opt.comp = list("tree", "multi", "aov", pvalue = 0.01))
```

